Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-05 paper.tex
reworked section on networks in drug discovery

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined
\cite{Imming:2006lt}, it has become apparent that this model does not
always match reality. Instead, it has become increasingly clear that
many drugs (and more generally, small molecules) exhibit
polypharmacological behavior - interacting with multiple targets,
including their putative primary target. In many cases, the activity
against the primary target is significantly higher than off-target
activities, minimizing the effects of the latter. However, this is not
always the case as in the example of terfenadine (a histamine H1
antagonist), where off target activity against hERG
\cite{Crumb:1995it} led to an increased risk of cardiac death. But
more importantly, it is becoming increasingly clear that many diseases
are not driven by dysfunction of a single a target. Examples include
diabetes and cancer. In parallel, even when a single target is the
main driver for a given disease state, compensatory pathways can
``route around'' inhibition of the primary target, thus restoring the
functional behavior that the primary target is responsible (i.e., the
system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. Finally, it should be noted
that while the preceding discussion presents polypharmacological
behavior as a negative characteristic, this is not always the
case. Indeed, modern approaches to drug development now consider such
behavior attractive leading to efforts to rationally design or control
polypharmacology \cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

@@

\section{The Role of Networks in Drug Discovery}  
\label{sec:role-networks-drug}
Omics technologies are enabling us today to elucidate and interrogate
complex relationships between disease, proteins, genetic material and
small molecules at an unprecedented speed and scale. Interrogation of
relationships reveal common mechanism of action. This may lead to a
therapeutic candidate finding application in a seemingly unrelated
disease (repositioning) that share a previously unknown causal
mechanism. It also allows identification of drug pairs that show
synergistic activity stemming from their interactions with
complementary pathways. This type of relationship elucidation may also
prevent unwanted side-effects if interrogation would reveal an
unexpected link to a vital pathway that ought to remain
unperturbed. The process of interrogation starts with experiments
aimed at elucidating molecular mechanisms and pathways such as yeast
two-hybrid techniques, protein and small molecule mass spectrometry,
microarray and copy number variation studies. Experiments may also
include high throughput screening of millions of chemical compounds
against specific biological targets of therapeutic interest to find
respective agonists and antagonists. These experiments result in
pairwise interaction relationships from which the functional big
picture biological network can then be elucidated
\cite{Wu2010}. Algorithms that detect communities from complex
networks eg. community structure algorithm \cite{girvan2002community}
may be used. These network analyses algorithms allow discovery of a
set of sparsely interconnected subsets of vertices where the subsets
themselves are however densely connected within. In the biological
context, this allows connecting a set of biological cycles and smaller
pathways into a big picture biological metacommunity network from
which inferences on cross-reactivity and new application areas of a
drug candidate can be drawn.

Network based approach to mechanism elucidation of small molecules
that allow identification of synergistic pairs have been implemented
in tools such as the DrugComboRanker
\cite{huang2014drugcomboranker}. The authors use a community based
partitioning to create a drug metacommunity network and a disease
signaling metacommunity network. Drug targets are then mapped onto
disease networks to allow identification of synergistic pairs of drugs
and complementary mechanisms of action.  In the tool MANTRA (Mode of
Action by NeTwoRk Analysis)\cite{iorio2010discovery}, authors are able
to deduce mechanisms of action of new molecules and suggest
repositioning of a well known molecule to a new area of application.

Network analysis has also found its application in medicinal chemistry
and high-throughput screening. Cheminformatics based chemical scaffold
and topology networks capture the structure-activity relationships
between compounds in screening experiments \cite{Varin:2011ve}. These
help in medicinal chemistry efforts towards the identification and
optimization of lead compounds for the biological target of interest
by allowing visualization of the chemical space in the backdrop of
biological activity.

@@

Since small molecules are fundamentally weighted networks (of atoms
and bonds), a variety of graph algorithms can be applied to them to
derive invariants (also known as topological descriptors
\cite{Guha:2012vn}). However, small molecules interact with protein
targets and the targets themselves interact with each other. Such
explicit interactions lend themselves naturally to network
representations. More generally, observed or computed relationships
between small molecules, protein or gene targets and diseases allow
one to develop network representations which can then be visualized
and quantified to support integrative analyses of multiple data
types. In the following subsections we highlight applications of this
approach and where possible provide examples of R code to generate and
analyze such networks.

>>>>

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined
\cite{Imming:2006lt}, it has become increasingly clear that many drugs
(and more generally, small molecules) exhibit polypharmacological
behavior - interacting with multiple targets, including their putative
primary target. In many cases, the activity against the primary target
is significantly higher than off-target activities, minimizing the
effects of the latter. However, this is not always the case as in the
example of terfenadine (a histamine H1 antagonist), where off target
activity against hERG \cite{Crumb:1995it} led to an increased risk of
cardiac death. Concurrently, it is becoming increasingly clear that
many diseases are not driven by dysfunction of a single
target. Examples include diabetes and cancer. Even when a single
target is the main driver for a given disease state, compensatory
pathways can ``route around'' inhibition of the primary target, thus
restoring the functional behavior that the primary target is
responsible (i.e., the system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. While the preceding discussion
presents polypharmacological behavior as a negative characteristic,
this is not always the case. Indeed, modern approaches to drug
development now consider such behavior attractive, leading to efforts
to rationally design or control polypharmacology
\cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

@@

\section{The Role of Networks in Drug Discovery}  
\label{sec:role-networks-drug}

Computational approaches in drug discovery, such as docking and
statistical QSAR, tend to focus on interactions between small
molecules and protein targets. Network representations and models play
a variety of roles in the \emph{in silico} drug discovery pipeline
ranging from visualization of heteregeneous datasets
\cite{Huan:2010cy,Kuhn:2010zr} to developing predictive models of drug
action \cite{Vina:2009oq,Folger:2011tx}. A necessary requirement for
network based analyses is the availability of network (or interaction)
data. This can come in many forms. Examples include protein-protein
interaction (PPI) databases such as BioGRID
\cite{Chatr-Aryamontri:2015yf} and STRING \cite{Franceschini:2013qa}
and drug-target databases such as DrugBank
\cite{Wishart:2008aa}. Bioactivity databases such as Pubchem and
ChEMBL \cite{Gaulton:2012jl} also provide useful information on
drug-target relationships, based on measured binding affinities.  In
addition to small molecules and proteins, databases such as PharmGKB
\cite{Thorn:2005aa} and MetaCore (a commercial product from Thompson
Reuters) integrate small molecules, protein and gene targets and the
published literature, providing a rich source of relationships between
these entities.

Using these databases, network methods can be applied to various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness may lead to significant side effects in other parts of
the system [REF] and can be the source of observed side effects. On
these lines, approaches to characterize toxicity have employed network
analyses \cite{Kim:2014xs,Zhang:2014hj}. While some approaches have
considered side effects [REF] others have considered the problem at
the genomic level by characterizing gene networks that are perturbed
by small molecules. Campaigns focused on combination therapies can
also benefit from a network representation and network inference
approaches \cite{Huang:2014by,Tang:2013lr} to identify target
combinations whose simultaneous inhibition leads to synergistic
effects. Mechanism of action studies can benefit from network analyses
by integrating multiple data types. An example of this was described
by Young et al \cite{Young:2008aa} that integrated high content
screening and ligand-target prediction models to construct network
representations of MoA hypotheses. Network approaches have also been
proposed for SAR analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at
the lead optimization stage of the discovery pipeline. Drug
repositioning efforts \cite{Dudley:2011dn,Gottlieb:2011rx} benefit
from network analyses such as community detection algorithms
\cite{girvan2002community}.

@@

Since small molecules are fundamentally weighted networks (of atoms
and bonds), a variety of graph algorithms can be applied to them to
derive invariants (also known as topological descriptors
\cite{Guha:2012vn}). However, small molecules interact with protein
targets and the targets themselves interact with each other. Such
explicit interactions lend themselves naturally to network
representations. More generally, observed or computed relationships
between small molecules, protein or gene targets and diseases allow
one to develop network representations of these entities which can
then be visualized and quantified to support integrative analyses of
multiple data types. In the following subsections we highlight
applications of this approach and where possible provide examples of R
code to generate and analyze such networks.
